---
title: "Digoxin Model"
author: Ben Ridenhour
date:  "`r format(Sys.Date(), '%d-%b-%y')`"
output: 
  html_notebook:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## A Simple Analysis of Digoxin Based on Weight and Time

Given a data set containing 4 variables (subject ID, visit date, weight, and the mM concentration of digoxin), we have been asked to create a simple baseline metabolic model for digoxin in the blood. We have been told two facts: First, digoxin changes proportionally to changes in weight. Second, digoxin level are known fluctuate in a roughly thirty day cycle. Details of our analysis are given below.

## Data Processing

First we read in the data and verify it.

```{r}

digoxin <- read.csv("Digoxin.csv", header = T)

str(digoxin)
```

By looking at the structure of the data using the `str` command, we see two problems. First, we must format the date properly. Second, weight---which should be numeric---is being treated as a factor. This quickly fixes the date variable:

```{r}
digoxin$Visit <- as.Date(digoxin$Visit, format = "%m/%d/%y")
```

Next, we need to fix the weight variable. First, look at the data frame. We'll use the `xtable` library to make things look nice. We need to load the library:

```{r}
library(xtable)
```

Then run:
```{r results="asis"}
#print first 10 lines only
print(xtable(head(digoxin,10), align = c("l","c","c","r","r")),
      type="html", include.rownames=FALSE)
```

Clearly the first row has an improperly formatted weight, so we have to fix that.

```{r}
#First change a factor to a character vector, then a numeric one
digoxin$Weight <- as.numeric(as.character(digoxin$Weight))
#Next, the first element of the vector is now NA due to the improper formating; fix this by
#inserting the proper value
digoxin$Weight[1] <- 70
#Finally, check the structure again:
str(digoxin)
```

Our data now look good!

## Plot the Data

Before analyzing the data, let's plot it to get a feel for it. We'll use `ggplot2` for this task.

```{r echo=FALSE}
library(ggplot2)

g <- ggplot(data=digoxin, aes(x=Visit, y=mM_Digoxin, group = Subject))

#note plus sign at end of line to add a linebreak in the ggplot command
g + geom_line(aes(color = Subject)) + theme_bw() + ylab("mM digoxin") + xlab("Visit (time)") +
   ggtitle("Digoxin Levels over Time")

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot. When plotted, our data look as we would expect them and we can see both linear and cyclical trends. The next step is running some models.

## Some Possible Models

We were given to facts about our data that indicate that 

$$ y = a + b_1 x_1 + b_2 sin(\frac{2 \pi}{30} x_2)$$

is a reasonable model for digoxin, where $y = [digoxin]$, $x_1 = weight$, and $x_2=time$. We will run two versions of the model. In the first version, we estimate the parameter $a$; in the second version, $a=0$.

```{r}
m1 <- lm(mM_Digoxin ~ Weight + I(sin(2*pi/30*as.numeric(Visit))), data=digoxin)
summary(m1)

m2 <- lm(mM_Digoxin ~ Weight + I(sin(2*pi/30*as.numeric(Visit))) - 1, data=digoxin)
summary(m2)
```

### Comparing models using ANOVA

We can ask R which of our two models might be a better choice using the `ANOVA` function. Let's do that and then plot the diagnostics of the better model.

```{r}
anova(m1,m2) #m2 would be favored
plot(m2)
```

The ANOVA tells us the model without an intercept is better; the diagnostics for this model look good. 